---
title: "Lab 3 - Data Types | Exploring Datasets"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Spring 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Date objects

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] #selects the 3rd entry in vector1, NOTENOTENOTE: R starts at 1, Python
#starts at 0. Don't get confused!

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) #creates matrix with values b/w 1-20
matrix1 #NOTICE values are listed top-to-bottom, then across, based con nrow &
#ncol
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #this produces a matrix
#with the same values as above, but lists values left-to-right, then down
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
<<<<<<< HEAD
                                  c("un", "deux", "trois", "cat"))) #same as
#as previous (matrix2), but with specific row and column names. NOTICE, in dim-
#names = row names list first, then column names 

matrix1[4, ] #returns the 4th (penultimate) row of matrix1
matrix1[ , 3] #returns the 3rd column of matrix1
matrix1[c(12, 14)] #returns the values in the 12th & 14th position in the matrix
#reads the matrix the way the default write a matrix (top-to-bottom, then across)
matrix1[c(12:14)] #returns values in the 12th THROUGH 14th position
matrix1[2:4, 1:3] #returns a 2D subset of matrix. In this case columns 1-3 for 
#rows 2-4
=======
                                  c("un", "deux", "trois", "cat"))) #
matrix3

matrix1[4, ] #
matrix1[ , 3] #
matrix1[c(12, 14)] #
matrix1[c(12:14)] #
matrix1[2:4, 1:3] #
>>>>>>> b3c7106116c07efb5453f2e1b6ba3c74dfab9fdd

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) #combines the previous 3 lines to create a
#highly customized matrix. NOTICE specific values, writing direction ("byrow = 
#TRUE"), number of rows and columns, & names of rows and columns
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]] #this shit is cool b/c it shows that lists can contain objects within
#them with different datatypes ITS THE SAME AS a Python dictionary (keys & values)

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? num
e <- c("red", "white", "red", NA) # What type of vector? chr
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logi
dataframe1 <- data.frame(d,e,f) #
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) #this does a 
# couple things:
#   renames fields in dataframe using c()
#   connects two actions in one line using ';'

dataframe1[1:2,] #displays rows 1-2 for all fields
dataframe1[c("ID","Passed")] #displays all values for fields 'ID' & 'Passed' 
#^appears as smaller df
dataframe1$ID #displays one field of df as vector, couple probably '<-' it to 
#its own object


```
Question: How do the different types of data appear in the Environment tab?

> Answer: vectors appear as 'Values' while dataframes, matrices, & lists appear as 'Data'. What is the difference and why does it matter?

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: When I do that and hit tab it selects the 1st field in the dataframe. At least that's what happens when I typed it all continuously. If I type 'dataframe1$' and then hit tab, but then remove 'ID' (the 1st field), and THEN hit tab again, it brings the list of all fields up again. Useful for selecting fields.

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
high_tempDurham <- c(58, 41, 43, 45, 35, 42, 49, 52, 56, 66) #you can add 'L' to
# the back of each number values to specify a long integer, you can also specify
# a short integer with 'S' (maybe test)
low_tempDurham <- c(31, 20, 25, 28, 19, 23, 25, 29, 43, 49)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius. Use a function to create the two new vectors from your existing ones in Fahrenheit.

```{r}
F_to_C_temps <- function(x){
  (x-32)*(5/9)
}
# F_to_C_temps <- function(x){
#   round((x-32)*(5/9), 2)
# }
#^this function incorporates a final rounding into the conversion, useful for 
#cleaning df's
high_tempC <- F_to_C_temps(high_tempDurham)
low_tempC <- F_to_C_temps(low_tempDurham)
```

Combine your four vectors into a data frame and add informative column names.

```{r}
Durham10DayTemps <- data.frame("High F" = high_tempDurham, "Low F" = low_tempDurham, "High C" = high_tempC, "Low C" =low_tempC) #set field names using the format above, set
#row names using "row.name =" 
##Durham10DayTemps
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}
summary(Durham10DayTemps)
sd(Durham10DayTemps$Low.F) #sd seems to only work for one field at a time
#could be brute-forced, but maybe avoid that
summary(Durham10DayTemps$Low.F)
```

### Date objects

Remember formatting of dates in R: 

%d  day as number (0-31)
%m  month (00-12, can be e.g., 01 or 1)
%y  2-digit year
%Y  4-digit year
%a  abbreviated weekday
%A  unabbreviated weekday
%b  abbreviated month
%B  unabbreviated month

```{r}
# Adjust date formatting for today
# Write code for three different date formats. 
# An example is provided to get you started.
# (code must be uncommented) -----> what does this mean? How to I version control
#in GitHub?
today <- Sys.Date()
format(today, format = "%B")
format(today, format = "%y")
format(today, format = "%m-%d-%Y")
#format(today, format = "")

```


### Package lubridate

Install and load the package lubridate into your R session. Lubridate offers fast and user friendly parsing of date-time data. Create a string for today's data and then convert it to R date object using lubridate.

More info on lubridate [here][https://cran.r-project.org/web/packages/lubridate/lubridate.pdf].

```{r DateExercise1}
#install.packages("lubridate")
library(lubridate)

#Ex1
str_today <- "2021-feb-3"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#Ex2
str_today <- "21-feb-3"
#Since the format is year-month-day we will use function ymd()
date_obj_today <- ymd(str_today)
date_obj_today

#there are other similar functions ydm(), mdy(), etc
```

```{r DateExercise2}
#century issue
str_past <- "55-feb-3"
date_obj_past <- ymd(str_past)
date_obj_past

#Build a function to fix year that is more general than the one discussed in the lesson
fix.early.dates <- function(d, cutoff) {
       m <- year(d) %% 100  #operator %% is a modular division i.e. integer-divide year(d) by 100 and returns the remainder 
       year(d) <- ifelse(m > cutoff, 1900+m, 2000+m)  #this will update year(d), year() is a function that returns the year for a data object
       return(d) #this line may be unnecessary
}#NOTICE this function reguires 2 inputs: d -> the date & cutoff -> in this case 21 for 2021 (maybe)
#should be 22 now...

fixed_date_obj_past <- fix.early.dates(date_obj_past,cutoff=21) #cutoff could be the current year to be more general or any other depending on data set 
fixed_date_obj_past
```

```{r centuryfix}
#Fix for century issue #SOOOOOO much easier than the example in previous chunk
str_past <- "55-feb-3"
#Alternative 1
date_obj_past <- fast_strptime(str_past,"%y-%b-%d",cutoff_2000=22L)
date_obj_past

#Alternative 2
date_obj_past2 <- parse_date_time2(str_past,"ymd",cutoff_2000=22L)
date_obj_past2

#Functions ymd(), mdy(), ydm() do not take argument cutoff_2000
```

In some cases when dates are provided as integers, you may need to provide an origin for your dates. For example, excel date could be given as number of days since an origin date. Origin date can be different. When R looks at dates as integers, its origin is January 1, 1970. Check if that is true on your machine. 

```{r dateorigin}
#Check if "1970-01-01" is your origin date.
lubridate::origin
#I thiink that is you are converting Integer dates to current dates you may need
#to write a function, but I should check lubridate docs to see if it has a better
#fix
```