---
title: "Assignment 5: Data Visualization"
author: "Cal Oakley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r, message = FALSE, warning = FALSE}
#1
getwd()
library(tidyverse) #for plotting 
library(cowplot) #for plotting multiple plots simultaneously 
library(lubridate) #for reformatting dates later
PeterPaul_ChemNut <- read.csv(
  "../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
NiwotLitter <- read.csv(
  "../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv"
) #loading datasets
#2 
str(PeterPaul_ChemNut) #to check if dates are read as date format
str(NiwotLitter)
PeterPaul_ChemNut$sampledate  <- ymd(PeterPaul_ChemNut$sampledate) 
NiwotLitter$collectDate <- ymd(NiwotLitter$collectDate) #reformatting dates
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
CalsTheme <- theme_classic(base_size = 16) +
  theme(axis.text = element_text(color = "gray"), legend.position = "left",
        legend.justification = 2)
theme_set(CalsTheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r, warning = FALSE, message = FALSE}
#4
A05_plot1 <- ggplot(PeterPaul_ChemNut, aes(x = tp_ug, y = po4,
                                           color = lakename)) +
  geom_point() +
  ylim(0, 50) +
  labs(x = "Total Phosphorus", y = "PO4", color = "Lake Name") +
  geom_smooth(method = lm, color = "black")
print(A05_plot1)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r, warning = FALSE, message = FALSE}
#5
A05_Plot2 <- ggplot(PeterPaul_ChemNut, aes(x = month, y = temperature_C,
                    fill = lakename)) +
  geom_boxplot() +
  labs(fill = "Lake Name", x = "Month", y = "Temperature C") +
  theme(legend.key.size = unit(0.1, "cm"), 
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10))

A05_Plot3 <- ggplot(PeterPaul_ChemNut, aes(x = month, y = tp_ug,
                    fill = lakename)) +
  geom_boxplot() +
  ylim(0,100) +
  labs(x = "Month", y = "Total P")

A05_Plot4 <- ggplot(PeterPaul_ChemNut, aes(x = month, y = tn_ug,
                    fill = lakename)) +
  geom_boxplot() +
  ylim(0, 2000)
  labs(x = "Month", y = "Total N")

A05_cowplot <- plot_grid(A05_Plot2,
                         A05_Plot3 + theme(legend.position = "none"),
          A05_Plot4 + theme(legend.position = "none"),
          #+ theme(legend.position = "none") removes redunant legends from plot
          ncol = 3, align = 'h', rel_widths = c(.4, .25, .25))
print(A05_cowplot)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperatures in Paul Lake during early Summer are essentially the same as the temperatures in Peter Lake in late Summer/early Fall. Meanwhile the total Phosphorus and Nitrogen increase in Peter Lake later in the summer when compared with Paul Lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r, warning = FALSE, message = FALSE}
#6
# A05_plot5 <- ggplot(subset(NiwotLitter, format.Date(collectDate, "%m") >= 06 &
#                              format.Date(collectDate, "%m") <= 09,
#                            functionalGroup == "Needles")) +
#                       geom_point(aes(x = collectDate, y = subplotID,
#                                      color = nlcdClass,
#                                      shape = nlcdClass)) +
#   scale_x_date(date_breaks = "1 month", date_labels = "%m-%Y") +
#   theme(axis.text.x = element_text(angle = 45,  hjust = 1))
#above is my first attempt at solving this, I realized that there were many
#months for which data didn't exist, but I couldn't find a way to subset dates
#by just month. Each time I did this it said that it couldn't find my y variable.
#My subsetting must be confusing ggplot into thinking there is no data
#I settled for the code below because I think it meets the requirements
A05_plot5 <- ggplot(subset(NiwotLitter, functionalGroup == "Needles")) +
                      geom_point(aes(x = collectDate, y = dryMass,
                                     color = nlcdClass,
                                     shape = nlcdClass)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = "Year", y = "Dry Mass (g)", color = 'NLCD Class',
       shape = 'NLCD Class') +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1),
        legend.key.size = unit(0.1, "cm"), 
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10))
print(A05_plot5)
#7
A05_plot6 <- ggplot(subset(NiwotLitter, functionalGroup == "Needles")) +
                      geom_point(aes(x = collectDate, y = dryMass)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = "Year", y = "Dry Mass (g)", color = 'Year') +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1)) +
  facet_wrap(vars(nlcdClass, ncol = 3))
print(A05_plot6)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Originally I thought 6 was more effective because 7 didn't display the dates as well, but then I tried knitting and realized 6 didn't deal with dates much better. So I think 7 is better, because although it has the same issues with dates (where their ticks are positioned and their poor resolution) that 6 does, it is very clear which samples are from which NLCD Classes.