---
title: "Assignment 09: Data Scraping"
author: "Cal Oakley"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

# Total points:

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_09_Data_Scraping.Rmd") prior to submission.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE, warning=FALSE}
#1
getwd()
library(tidyverse)
library(rvest)
library(lubridate)

CalsTheme <- theme_classic(base_size = 16) +
  theme(axis.text = element_text(color = "gray"), legend.position = "left",
        legend.justification = 2)
theme_set(CalsTheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2019 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Change the date from 2020 to 2019 in the upper right corner.
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020>

Indicate this website as the as the URL to be scraped. (In other words, read the contents into an
`rvest` webpage object.)

```{r set.the.scraping.website, message = FALSE, warning=FALSE}
#2
theWebsite <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2020')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Average Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to three separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values, with the first value being 36.0100.

```{r scrape.the.data, message = FALSE, warning=FALSE}
#3
water.system.name <- theWebsite %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
pwsid <- theWebsite %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
ownership <- theWebsite %>% 
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
max.withdrawals.mgd <- theWebsite %>% 
  html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in order. You can overcome this by creating a month column in the same order the data are scraped: Jan, May, Sept, Feb, etc...

5. Plot the max daily withdrawals across the months for 2020

```{r create.a.dataframe.from.scraped.data, message = FALSE, warning=FALSE}
#4
DurhamH2O_2020 <- data.frame(
  "Month" = c('Jan','May','Sep','Feb','Jun','Oct',
              'Mar','Jul','Nov','Apr','Aug','Dec'),
  "Year" = rep(2020, 12),
  "WaterSystemName" = water.system.name,
  "PWSID" = pwsid,
  "Ownership" = ownership,
  "MaxDailyWthdrl" = as.numeric(max.withdrawals.mgd)) %>% 
  mutate(Date = my(paste(Month, '-', Year))) %>% 
  arrange(-desc(Date), .by_group = FALSE)

head(DurhamH2O_2020)

#5
A09_plot1 <- ggplot(DurhamH2O_2020, aes(x=Date, y=MaxDailyWthdrl)) +
  geom_line() +
  geom_smooth(method = 'loess', se = FALSE) +
  labs(title = paste("2020 Water usage data for",water.system.name,
                     ownership),
       subtitle = pwsid,
       y="Max Withdrawal (mgd)",
       x="Date")

print(A09_plot1)

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site scraped**.

```{r construct.a.scraping.function, message = FALSE, warning=FALSE}
#6.
scraper <- function(thePWSID,theYear){
  theWebsite <- read_html(
    paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
           thePWSID,'&year=',theYear))
  
  water.system.name <- theWebsite %>% 
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text()
  pwsid <- theWebsite %>% 
    html_nodes('td tr:nth-child(1) td:nth-child(5)') %>% html_text()
  ownership <- theWebsite %>% 
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text()
  max.withdrawals.mgd <- theWebsite %>% 
    html_nodes('th~ td+ td') %>% html_text()
  
  df <- data.frame(
    "Month" = c('Jan','May','Sep','Feb','Jun','Oct',
                'Mar','Jul','Nov','Apr','Aug','Dec'),
    "Year" = rep(theYear, 12),
    "WaterSystemName" = water.system.name,
    "PWSID" = pwsid,
    "Ownership" = ownership,
    "MaxDailyWthdrl" = as.numeric(max.withdrawals.mgd)) %>% 
      mutate(Date = my(paste(Month, '-', Year))) %>% 
      arrange(-desc(Date), .by_group = FALSE)
  
  return(df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, message = FALSE, warning=FALSE}
#7
q7_df <- scraper('03-32-010','2015')

head(q7_df)
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares the Asheville to Durham's water withdrawals.

```{r fetch.and.plot.AshevilleDurham.2015.data, message = FALSE, warning=FALSE}
#8
q8_df <- scraper('01-11-010','2015') %>% 
  bind_rows(q7_df)

A09_plot2 <- ggplot(q8_df, aes(x=Date, y=MaxDailyWthdrl,
                               color = WaterSystemName)) +
  geom_line() +
  labs(title = '2015 Water usage data',
       y="Max Withdrawal (mgd)",
       x="Date",
       color='') +
  theme(legend.position = 'right')
print(A09_plot2)
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

```{r, message = FALSE, warning=FALSE}
#9
q9_years <- as.character(c(2010:2019))

q9_df <- map2('01-11-010', q9_years, scraper) %>% bind_rows()

A09_plot3 <- ggplot(q9_df, aes(x=Date, y=MaxDailyWthdrl)) +
  geom_line() +
  geom_smooth(method = 'loess', se = FALSE) +
  labs(title = '2015 Water usage data for Asheville Municipality',
       subtitle = '01-11-010',
       y="Max Withdrawal (mgd)") +
  theme(legend.position = 'right')
print(A09_plot3)
```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

>It does appear that Asheville has increased its water usage over the last 10 years.